Rectangulotopes

### 2024-04-22 draft/rectangulotopes.tex
minor changes

\begin{figure}
  \includegraphics[width=.9\textwidth]{rectangulations.pdf}
	\caption{Three rectangulations. All three rectangulations are weakly equivalent, but only the first two are strongly equivalent. Only the third one is diagonal.}
	\label{fig:rectequiv}
\end{figure}
\vincent{Add one which is not weakly equivalent?}

@@

\begin{theorem}
  \label{thm:weakRectangulotope}
  % Jean: I chose to denote the *abstract* polytope by WRP(n), hence changed the next sentence accordingly.
%The weak rectangulotope is realized by the polytope~$\WRP (n) \subset \R^n\!$ defined equivalently~as
The weak rectangulotope $\WRP (n)$ is realized by: \vincent{We loose the information that the two descriptions define the same polytope...}
  \begin{enumerate}[(i)]
  \item the convex hull of the points
    \[
    \sum_{i\in [n]} (\loday{w}^R_i - \antiloday{w}^R_i)\cdot \b{e}_i
    \]
    for all weak rectangulations $R$ of size~$n$, with
  \[
    \loday{w}^R_i \eqdef h^T_i\cdot v^T_i
    \qquad\text{and}\qquad
    \antiloday{w}^R_i \eqdef h^S_i\cdot v^S_i,
  \]
  where~$S$ and~$T$ are the source and target trees of the rectangulation~$R$, and $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in S} x_i \le \#\set{I \text{ interval of } [n]}{I \not\subseteq S \text{ and } I \not\subseteq [n] \ssm S}
  \]
  for all subsets~$\varnothing \ne S \subsetneq [n]$.
  \end{enumerate}
\end{theorem}

@@

\begin{comment}
\begin{example}
  The vertex of~$\WRP (16)$ corresponding to the weak rectangulation of \cref{fig:weakRectangulation} is
  \[
  (-3, 0, 26, 2, -9, 5, -69, -4, 17, 0, -8, 59, 2, -20, 0, 2).
  \]
  For instance, the $4$th coordinate is $h^T_4 \cdot v^T_4 - h^S_4 \cdot v^S_4 = 1 \cdot 5 - 1 \cdot 3 = 2$.
  The facet of~$\WRP (16)$ corresponding to the subset~$\{3,4,7,8,9,10,14\}$ is given by the inequality
  \[
  x_3 + x_4 + x_7 + x_8 + x_9 + x_{10} + x_{14} \le ???
  \]
  \vincent{Compute and check...}
\end{example}
\end{comment}
\vincent{not sure if we want to reintroduce this example.}

@@

\begin{theorem}
\label{thm:strongRectangulotope}
% The strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
The strong rectangulotope~$\SRP (n)$ is realized by:
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ of size~$n$, with
  \[
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$,
  \end{itemize}
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}
  \]
  for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ denote nonempty and consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

\begin{comment}
\begin{example}
  The vertex of~$\SRP (16)$ corresponding to the strong rectangulation of \cref{fig:strongRectangulation} is
  \[
  (-41, 5, 262, 24, -110, 50, -525, -50, 184, 3, -88, 450, 35, -221, -5, 27).
  \]
  For instance, the contribution of~$\b{e}_5 - \b{e}_7$ is ...
  \vincent{Compute and check...}
  The facet of~$\SRP (16)$ corresponding to the subset~$\{3,4,7,8,9,10,14\}$ is given by the inequality
  \[
  x_3 + x_4 + x_7 + x_8 + x_9 + x_{10} + x_{14} \le ???
  \]
  \vincent{Compute and check...}
\end{example}
\end{comment}

@@

\pagebreak
\section{Weak rectangulotopes}
\label{sec:weakRectangulotopes}

>>>>

\begin{figure}
  \includegraphics[width=.65\textwidth]{rectangulations.pdf}
	\caption{Three rectangulations. All three rectangulations are weakly equivalent, but only the first two are strongly equivalent. Only the third one is diagonal.}
	\label{fig:rectequiv}
\end{figure}
\vincent{Add one which is not weakly equivalent?}

@@

\begin{theorem}
  \label{thm:weakRectangulotope}
  % Jean: I chose to denote the *abstract* polytope by WRP(n), hence changed the next sentence accordingly.
%The weak rectangulotope is realized by the polytope~$\WRP (n) \subset \R^n\!$ defined equivalently~as
The weak rectangulotope $\WRP (n)$ is realized by the polytope equivalently described as:
  \begin{enumerate}[(i)]
  \item the convex hull of the points
    \[
    \sum_{i\in [n]} (\loday{w}^R_i - \antiloday{w}^R_i)\cdot \b{e}_i
    \]
    for all weak rectangulations $R$ of size~$n$, with
  \[
    \loday{w}^R_i \eqdef h^T_i\cdot v^T_i
    \qquad\text{and}\qquad
    \antiloday{w}^R_i \eqdef h^S_i\cdot v^S_i,
  \]
  where~$S$ and~$T$ are the source and target trees of the rectangulation~$R$, and $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in S} x_i \le \#\set{I \text{ interval of } [n]}{I \not\subseteq S \text{ and } I \not\subseteq [n] \ssm S}
  \]
  for all subsets~$\varnothing \ne S \subsetneq [n]$.
  \end{enumerate}
\end{theorem}

@@

%\begin{example}
  The vertex of~$\WRP (16)$ corresponding to the weak rectangulation of \cref{fig:weakRectangulation} is
  \[
  (-3, 0, 26, 2, -9, 5, -69, -4, 17, 0, -8, 59, 2, -20, 0, 2).
  \]
  For instance, the $4$th coordinate is $h^T_4 \cdot v^T_4 - h^S_4 \cdot v^S_4 = 1 \cdot 5 - 1 \cdot 3 = 2$.
%  The facet of~$\WRP (16)$ corresponding to the subset~$\{3,4,7,8,9,10,14\}$ is given by the inequality
%  \[
%  x_3 + x_4 + x_7 + x_8 + x_9 + x_{10} + x_{14} \le ???
%  \]
%  \vincent{Compute and check...}
%\end{example}

@@

\begin{theorem}
\label{thm:strongRectangulotope}
% The strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
The strong rectangulotope~$\SRP (n)$ is realized by the polytope equivalently described as:
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ of size~$n$, with
  \[
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$,
  \end{itemize}
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}
  \]
  for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ denote nonempty and consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

%\begin{comment}
%\begin{example}
  The vertex of~$\SRP (16)$ corresponding to the strong rectangulation of \cref{fig:strongRectangulation} is
  \[
  (-41, 5, 262, 24, -110, 50, -525, -50, 184, 3, -88, 450, 35, -221, -5, 27).
  \]
%  For instance, the contribution of~$\b{e}_5 - \b{e}_7$ is ...
%  \vincent{Compute and check...}
%  The facet of~$\SRP (16)$ corresponding to the subset~$\{3,4,7,8,9,10,14\}$ is given by the inequality
%  \[
%  x_3 + x_4 + x_7 + x_8 + x_9 + x_{10} + x_{14} \le ???
%  \]
%  \vincent{Compute and check...}
%\end{example}
%\end{comment}

@@

\section{Weak rectangulotopes}
\label{sec:weakRectangulotopes}
